---
title: "MSstatsPTM for statistical quantification of PTMs"
author: "Tsung-Heng Tsai (<tsai.tsungheng@gmail.com>)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MSstatsPTM for statistical quantification of PTMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# devtools::load_all()
library(MSstatsPTM)
```

# Introduction

The goal of MSstatsPTM is to provide the implementation of general statistical methods for characterization of quantitative changes in global post-translational modification profiling experiments.

The MSstatsPTM package provides three categories of important functions for quantitative analysis of PTMs: summarization, estimation and comparison. 

- The `PTMsummarize` and related functions summarize log2-intensities of spectral features for each PTM site into one value per run.
- The `PTMestimate` and related functions take as input the summarized log2-intensities for each PTM site, performs statistical modeling for the abundance of the site, and returns the estimates of model parameters for all sites in all experimental conditions.
- The `PTMcompareMeans` and related functions perform significance analysis for detecting changes in PTM mean abundances between conditions.

# Required data format

```{r}
PTMsimulateExperiment(nGroup = 2, nRep = 2, nProtein = 1, nSite = 2, nFeature = 5,
                      list(PTM = 25, Protein = 25), list(PTM = c(0, 1), Protein = c(0, 1)),
                      list(PTM = 0.2, Protein = 0.2), list(PTM = 0.05, Protein = 0.05))
```

## Site-level representation

### Working with FASTA files

## Simulated data

# PTM data processing

## Normalization

### Normalization by equalizing medians

### Normalization with reference

# PTM summarization

## Summarization options

# PTM estimation

## Site-specific analysis

### Total protein abundance

## Data acquired in multiple batches

### Linear models with a fixed batch effect

### Separate linear models for different batches

# PTM comparison

## Detection of changes in mean abundances of PTM sites

## Protein-level adjustment

## Integration of inferences from multiple batches

